---
title: "Assignment1 Q9"
author: "Jiahang Wang"
output: pdf_document
---
```{r message=FALSE}
# Load necessary libraries
library(readr)
library(dplyr)
```


```{r message=FALSE}
# load data
data.source <- paste0("https://raw.githubusercontent.com/",
"mcgillstat/regression/main/data/medals_2008.csv")
medals <- read_csv(data.source, col_names = FALSE)
colnames(medals) <- c("Country", "Gold", "Silver", "Bronze")
# e.g. print out the first 11 countries
print(medals, n = 11)
```
# Q1
```{r}
# Calculate the number of nations with at least one gold medal
nations_with_gold <- medals %>%
  filter(Gold > 0) %>%
  nrow()

# Print the result
print(nations_with_gold)
```
# Q2
```{r}
# Center and scale the medal counts, keeping Country as the first column
medals_p <- medals %>%
  mutate(across(c(Gold, Silver, Bronze), scale)) 

medals_p
```
- Centering:

Centering is essential when data doesn't naturally have a mean of zero. If the data is not centered, the first principal component may capture the data's overall average instead of its underlying patterns and variances.
It's also necessary when variables have different means. Failing to center can lead to the PCA being sensitive to variables with larger means, potentially distorting the interpretation of principal components.
Centering ensures that the origin (0,0) in the multidimensional space corresponds to the center of data distribution, making it easier to interpret the directions of maximum variance.

- Scaling:

Scaling is important when variables have different units or scales (e.g., one variable is measured in meters, another in grams). Without scaling, variables with larger numeric ranges can dominate the PCA analysis, as PCA is based on variance, and larger values contribute more to variance.
By scaling the data, it will give equal importance to each variable, ensuring that the PCA components are influenced by the relative variations in each variable rather than their absolute values.
Scaling also helps in making the analysis less sensitive to the choice of units in which variables are measured, making the results more robust and interpretable.

# Q3
```{r}
# Perform Principal Component Analysis
pca_result <- prcomp(medals_p[, c("Gold", "Silver", "Bronze")])

# Extract the loadings (principal components)
loadings <- pca_result$rotation[, 1:3]

# Extract the eigenvalues
eigenvalues <- pca_result$sdev[1:3]^2

# Display the principal components
print("Principal Components (PC1, PC2, PC3):")
print(loadings)

# Display their corresponding eigenvalues
print("Corresponding Eigenvalues:")
print(eigenvalues)
```
# Q4
## a
```{r}
# Perform Principal Component Analysis
pca_result <- prcomp(medals_p[, c("Gold", "Silver", "Bronze")])

# Calculate the amount of variance captured by the first principal component
variance_captured_pc1 <- pca_result$sdev[1]^2

# Print the amount of variance captured by PC1
print(paste("Amount of variance captured by PC1:", variance_captured_pc1))

# Calculate the total variance of the dataset
total_variance <- sum(pca_result$sdev^2)

# Calculate the reconstruction error in terms of variance
reconstruction_error_variance <- total_variance - variance_captured_pc1

# Print the reconstruction error in terms of variance
print(paste("Reconstruction error:", reconstruction_error_variance))
```
## b
```{r}
# Perform Principal Component Analysis
pca_result <- prcomp(medals_p[, c("Gold", "Silver", "Bronze")])

# Calculate the amount of variance captured by the first two principal components
variance_captured_pc1_pc2 <- sum(pca_result$sdev[1:2]^2)

# Print the amount of variance captured by PC1 and PC2
print(paste("Amount of variance captured by PC1 and PC2:", variance_captured_pc1_pc2))

# Calculate the total variance of the dataset
total_variance <- sum(pca_result$sdev^2)

# Calculate the reconstruction error in terms of variance
reconstruction_error_variance <- total_variance - variance_captured_pc1_pc2

# Print the reconstruction error in terms of variance
print(paste("Reconstruction error:", reconstruction_error_variance))
```
## c
```{r}
# Perform Principal Component Analysis
pca_result <- prcomp(medals_p[, c("Gold", "Silver", "Bronze")])

# Calculate the amount of variance captured by the first three principal components
variance_captured_pc1_pc2_pc3 <- sum(pca_result$sdev[1:3]^2)

# Print the amount of variance captured by PC1, PC2, and PC3
print(paste("Amount of variance captured by PC1, PC2, and PC3:", variance_captured_pc1_pc2_pc3))

# Calculate the total variance of the dataset
total_variance <- sum(pca_result$sdev^2)

# Calculate the reconstruction error in terms of variance
# For three components, this error should be very small or zero, as we are using all components
reconstruction_error_variance <- total_variance - variance_captured_pc1_pc2_pc3

# Print the reconstruction error in terms of variance
print(paste("Reconstruction error:", reconstruction_error_variance))
```
# Q5
```{r}
# Perform PCA on the scaled data to obtain the principal component vectors
pca_result_scaled <- prcomp(medals_p[, c("Gold", "Silver", "Bronze")])

# Extract the first principal component vector (b1)
b1 <- pca_result_scaled$rotation[, 1]

# Project the data onto b1
original_centered <- scale(medals_p[, c("Gold", "Silver", "Bronze")])
coor_b1 <- as.data.frame(original_centered %*% b1)

# Add country names to the coordinates table
coor_b1$Country <- medals$Country

# Rename the first column to "PC1_Coordinates"
names(coor_b1)[1] <- "PC1_Coordinates"

coor_b1
```

## rank of nations based on the PC1 coordinates
```{r}
# Rank nations based on the PC1 coordinates
ranked_nations <- coor_b1 %>%
  arrange(PC1_Coordinates)

# Print the ranked list of nations
head(ranked_nations)
```
## rank of nations based on  three ranking methods
```{r}
# Add overall rank to ranked_nations
ranked_nations$Rank_PCA1 <- seq_along(ranked_nations$PC1_Coordinates)

# Calculate ranks in the original medals data frame
medals$Rank_Gold = rank(-medals$Gold, ties.method = "min")
medals$Rank_Total = rank(-(medals$Gold + medals$Silver + medals$Bronze), ties.method = "min")

# Merge these ranks into the ranked_nations data frame and exclude the PC1_Coordinates column
ranked_nations_all <- ranked_nations %>%
  left_join(select(medals, Country, Rank_Gold, Rank_Total), by = "Country") %>%
  select(-PC1_Coordinates)

# Display the first few rows of the new data frame
head(ranked_nations_all, n=10)
```
## differences of three ranking methods
```{r}
# Create a new dataframe with differences
result_diff <- data.frame(
  Countries = ranked_nations_all$Country,
  Diff_PCA1_Gold = abs(ranked_nations_all$Rank_PCA1 - ranked_nations_all$Rank_Gold),
  Diff_PCA1_Total = abs(ranked_nations_all$Rank_PCA1 - ranked_nations_all$Rank_Total),
  Diff_Gold_Total = abs(ranked_nations_all$Rank_Gold - ranked_nations_all$Rank_Total)
)

# Display the first few rows of the new data frame
head(result_diff)

```
As shown above the ranking difference between 3 methods are both small. With the smallest difference between PCA1 ranking method and the total ranking method.


# Q6

Collect and normalize medal data for each nation. Create an autoencoder neural network architecture with an encoder to reduce the data dimensionality and a decoder to reconstruct the original data. Train the autoencoder using the normalized medal data and minimize the reconstruction error. And using the trained encoder to obtain lower-dimensional representations. Rank nations based on their positions in this reduced space, where nations closer together may have similar medal performances(rank). It is more capable of capturing non-linear relationships in the data, which PCA, being a linear method, cannot. And also automatically learn features, prioritizing data aspects more relevant for reconstruction, which can be more meaningful for analyzing complex medal data.





